Assessing United States county-level exposure for research on tropical cyclones and human health

### 2019-01-27 writing/ehp_results.tex
Rearrange content

\subsection*{Assessment of tropical cyclone exposure data}

Within
the nine sample counties, storm-related rainfall from the reanalysis precipitation dataset 
was well-correlated with station-based rainfall measurements, although the reanalysis data 
did tend to somewhat underestimate rainfall totals in storms with extremely high rainfall 
(Figure~\ref{fig:raincomparison}). A distance constraint (county population mean center within
500 kilometers of the storm's closest approach) was necessary for assessing storm exposure
based on this metric to ensure that rains unrelated and far from the storm track were not
misattributed to storm exposures. We found that this distance constraint was typically
adequately large to capture storm-related rain, although in rare examples of exceptionally
large storms (e.g., Hurricane Ike [2008]) or storms for which tracking was stopped at
extratropical transition (e.g., Tropical Storm Lee [2011]), some storm-related rains were
missed because of this distance constraint (Figure~\ref{fig:rainexamples}).

The modeled wind speeds generally agreed well
with the HURDAT2 wind radii (Figure~\ref{fig:windcomparison}), with disagreement mostly 
limited to large storms that were characterized by high winds well inland (e.g., Hurricanes
Ike [2008] and Sandy [2012]; Figure~\ref{fig:windexamples}).

Stream gage data generally indicated higher discharge in these
counties on dates identified as flood events, particularly in counties with more reporting
stream gages (e.g., Harris County, TX, with 10 reporting stream gages; 
Figure~\ref{fig:floodcomparison}), although these two flooding data sources were 
not completely
consistent.

@@

We first calculated summary statistics for the exposure assessments (Table
\ref{tab:exposuresummaries}). There was wide variation in the average number of
county exposures per year under the five different tropical cyclone metrics
considered. For the tornado metric, there were approximately 50 county
exposures identified per year, on average, within our study.  For flood and
wind metrics, there were substantially more county exposures identified (over
200 per year on average), and even more when assessing exposure using the
distance and rain metrics (over 400 per year on average).

We also measured the average number of county exposures per tropical cyclone,
within those tropical cyclones in which at least one county was exposed to the
hazard (Table~\ref{tab:exposuresummaries}). The typical number of county
exposures per tropical cyclone was lowest when assessing exposure using the
tornado metric (median: 7 county exposures per tropical cyclone), higher under
the flood and wind metrics (median: 26 and 34, respectively), and even higher for the metrics of rain (median: 68)
and distance (median: 84).  For every metric except the tornado-based metric,
we identified at least one tropical cyclone with over 300 counties exposed.
However, the largest-extent tropical cyclone varied by metric (e.g., Beryl in
1994 exposed the most counties based on distance, Frances in 2004 based on
rain, and Ike in 2008 based on wind).

@@

Figure~\ref{fig:ivanexposure} gives an example of how sensitive county-level
tropical cyclone exposure assessment is to the choice of metric, using as an
example Hurricane Ivan (2004), a record-breaking tropical cyclone for its
duration as a major hurricane~\citep{franklin2006atlantic}. When county-level
exposure was determined based on distance, the counties assessed as exposed
followed the tropical cyclone's track, including counties well inland. For the
wind metric, only counties near two of the tropical cyclone's landfalls were
assessed as exposed. For rain- and flood-based metrics, however, exposure
extended to the left of the track, including counties as far north as New York
and Connecticut, while for the tornado metric, exposed counties tended to be to
the right of the track and included several counties in central North Carolina,
South Carolina, and Georgia that were not identified as exposed to Ivan based
on any other metric.  Figure~S3 provides similar maps for four other example
tropical cyclones (selected because they exposed many \ac{US}  counties based on
at least one metric), and Figure~S4 provides four examples of tropical cyclones
in which the disagreement in exposure assessment between rain- and wind-based
metrics was particularly notable.

We drew similar conclusions when we investigated all 46 tropical cyclones
between 1996 and 2011 (the period for which all five metrics were available)
for which 250 or more counties were exposed based on at least one metric
(Figure~\ref{fig:jaccard}). In this figure, each row provides results for a
specific tropical cyclone, and each box in that row shows a measure of
agreement between a specific pair of metrics in terms of the counties assessed
as exposed to that tropical cyclone, with darker colors representing greater
agreement in county-level exposure assessments.

For most tropical cyclone, distance-based county-level exposure assessment was
at best in moderate agreement with exposure assessments based on the four
hazard-based metrics. For almost all tropical cyclones, Jaccard index values
between distance and any of the four hazard-based metrics were 0.6 or lower
(i.e., out of the counties assessed as exposed to the tropical cyclone by at
least one of the two metrics considered, 60\% or fewer of the counties had the
same exposure assessment under the two metrics). For many tropical cyclone, the
Jaccard index values were 0.2 or lower, suggesting very poor agreement in
exposure assessment between the metrics.  Among the four hazard-based metrics,
the tornado-based metric showed universally poor agreement with other metrics
in county-level classification across all tropical cyclones considered.  For
other pairs of metrics, there were also generally large differences in which
counties were determined to be exposed to the tropical cyclone, with the
Jaccard index values below 0.4 for most metric pairs in most tropical cyclones.

There were, however, a few exceptions\,---\,tropical cyclones in which similar
counties were determined to be exposed to the tropical cyclone for two or more
of the metrics considered.  For Floyd in 1999 (Figure~S3) and Irene in 2011,
for example, county-level classification agreed moderately to well (Jaccard
index of approximately 0.5\,--\,0.8) for all pairs of exposure metrics except those
including the tornado-based metric. These tropical cyclones both made their
first \ac{US}  landfall in North Carolina at minor hurricane windspeeds (Category
2 and 1, respectively) and then skimmed the eastern coast of the \ac{US} north
through New England, bringing large rainfall to much of the eastern coast from
North Carolina north and causing extensive inland flooding in North Carolina
(Floyd) and New England (Irene)~\citep{avila2013atlantic,
lawrence2000atlantic}. For another set of tropical cyclones (e.g., Lee in 2011
[Figure~S3], Ernesto in 2006, and Bertha in 1996), there was moderate to good
agreement for all pairwise combinations of distance, rain, and wind, but poor
agreement for all other combinations of metrics. Most of these tropical
cyclones either skimmed along the eastern coast north of North Carolina (Bertha
in 1996, Charley in 2004, Gaston in 2004, Barry in 2007, and Hanna in 2008), in
a pattern similar to Hurricanes Floyd in 1999 and Irene in 2011, or made
landfall in or near the Florida panhandle and cut across to exit into the
Atlantic around North Carolina (Josephine in 1996, Earl in 1998, Alberto in
2006). However, these two sets of tropical cyclones represented the minority of
all tropical cyclones considered; for most tropical cyclones, there was low
overlap in the counties determined to be exposed to the tropical cyclones
between most pairings of metrics.

@@

Strong geographical patterns were clear across tropical cyclone exposure
assessments based on these different metrics when the average number of
exposures per decade in each county were calculated 
(Figure~\ref{fig:averageexposure}). Wind-based exposure had a strong coastal pattern,
with almost all exposures in counties within about 200 kilometers (124 miles)
of the coastline, mostly in eastern North and South Carolina, southern Florida,
and the area near the outlet of the Mississippi River. While exposures by rain-
and distance-based metrics were also more common in coastal areas compared to
inland areas, inland exposures were much more common based on these two metrics
compared to the wind-based metric. Rain and, to some extent, flood exposures
were characterized by a pattern defined by the Appalachian Mountains, with much
lower exposures west of the mountain range than to the east. Almost all
tornado-based exposures were in coastal states, with most in Florida, Alabama,
South Carolina, and North Carolina and almost none north of Maryland.
Flood-based exposures were highest in North Carolina and along the South
Carolina coast, as well as further north in New Jersey, southeastern
Pennsylvania, and southeastern New York. Notably, some areas of the \ac{US} that
experienced regular exposures to tropical cyclone-related excessive rainfall
and floods were not identified as areas of regular risk based on the more
commonly used tropical cyclone exposure assessment metrics of wind or distance
(Figure~\ref{fig:averageexposure}).

We included a case study in the Supporting Information showing how these
differences in exposure assessment across the five metrics considered could
strongly influence health-related studies based on physical exposure to
tropical cyclones. For this example, we compared estimates of which \ac{US}
counties have the largest expected physical exposure to tropical cyclones among
a susceptible subpopulation, Medicare beneficiaries who are reliant on
electricity for medical equipment (Supplemental Text, Figure~S6)~\citep{empower}.

@@

We investigated how well these data correspond with data from other available
sources (Figures~S7\,--\,11).  While generally in agreement with data from other
sources, there are a few limitations. The wind data are based on modeled,
rather than observed, values, and while the modeled wind data generally agree
well with post-analysis maximum wind radii from the revised Atlantic hurricane
database (\ac{HURDAT2}~\citep{landsea2013}) (Figure~S7), for a few tropical cyclones
it did not (e.g., Figure~S10). The rainfall data are from re-analysis data,
which are generally well-correlated with observed ground-based station data but
sometimes oversmooth extreme measurements (Figure~S8). More details on the assessment
of this data are provided in the Supplemental Material.

>>>>

@@

We first calculated summary statistics for the exposure assessments
(Table~\ref{tab:exposuresummaries}). There was wide variation in the average
number of county exposures per year under the five different tropical cyclone
metrics considered. For the tornado metric, there were approximately~50 county
exposures identified per year, on average, within our study.  For flood and
wind metrics, there were substantially more county exposures identified ($>$200
per year on average), and even more when assessing exposure using the distance
and rain metrics ($>$400 per year on average).

We also measured the average number of county exposures per tropical cyclone,
within those tropical cyclones in which at least one county was exposed to the
hazard (Table~\ref{tab:exposuresummaries}). The typical number of county
exposures per tropical cyclone was lowest when assessing exposure using the
tornado metric (median:~7 county exposures per tropical cyclone), higher under
the flood and wind metrics (median:~26 and~34, respectively), and even higher
for the metrics of rain (median:~68) and distance (median:~84).  For every
metric except the tornado-based metric, we identified at least one tropical
cyclone with over~300 counties exposed.  However, the largest-extent tropical
cyclone varied by metric (e.g., Beryl in~1994 exposed the most counties based
on distance, Frances in~2004 based on rain, and Ike in~2008 based on wind).

@@

Figure~\ref{fig:ivanexposure} gives an example of how sensitive county-level
tropical cyclone exposure assessment is to the choice of metric, using as an
example Hurricane Ivan in 2004, which was record-breaking for its
duration as a major hurricane~\citep{franklin2006atlantic}. When county-level
exposure was determined based on distance, the counties assessed as exposed
followed the tropical cyclone's track, including counties well inland. For the
wind metric, only counties near two of the tropical cyclone's landfalls were
assessed as exposed. For rain- and flood-based metrics, however, exposure
extended to the left of the track, including counties as far north as New York
and Connecticut, while for the tornado metric, exposed counties tended to be to
the right of the track and included several counties in central North Carolina,
South Carolina, and Georgia that were not identified as exposed to Ivan based
on any other metric.  Figure~S3 provides similar maps for four other example
tropical cyclones (selected because they exposed many \ac{US}  counties based on
at least one metric), and Figure~S4 provides four examples of tropical cyclones
in which the disagreement in exposure assessment between rain- and wind-based
metrics was particularly notable.

We drew similar conclusions when we investigated all~46 tropical cyclones
between~1996 and~2011 (the period for which all five metrics were available)
for which~250 or more counties were exposed based on at least one metric
(Figure~\ref{fig:jaccard}). In this figure, each row provides results for a
specific tropical cyclone, and each box in that row shows a measure of
agreement between a specific pair of metrics in terms of the counties assessed
as exposed to that tropical cyclone, with darker colors representing greater
agreement in county-level exposure assessments.

For most tropical cyclone, distance-based county-level exposure assessment was
at best in moderate agreement with exposure assessments based on the four
hazard-based metrics. For almost all tropical cyclones, Jaccard index values
between distance and any of the four hazard-based metrics were~0.6 or lower
(i.e., out of the counties assessed as exposed to the tropical cyclone by at
least one of the two metrics considered,~60\si{\percent} or fewer of the
counties had the same exposure assessment under the two metrics). For many
tropical cyclones, the Jaccard index values were~0.2 or lower, suggesting very
poor agreement in exposure assessment between the metrics.  Among the four
hazard-based metrics, the tornado-based metric showed universally poor
agreement with other metrics in county-level classification across all tropical
cyclones considered.  For other pairs of metrics, there were also generally
large differences in which counties were determined to be exposed to the
tropical cyclone, with the Jaccard index values below~0.4 for most metric pairs
in most tropical cyclones.

There were, however, a few exceptions\,---\,tropical cyclones in which similar
counties were determined to be exposed to the tropical cyclone for two or more
of the metrics considered.  For Floyd in~1999 (Figure~S3) and Irene in~2011,
for example, county-level classification agreed moderately to well (Jaccard
index of approximately~0.5\,--\,0.8) for all pairs of exposure metrics except
those including the tornado-based metric. These tropical cyclones both made
their first \ac{US}  landfall in North Carolina at minor hurricane windspeeds
(Category~2 and~1, respectively) and then skimmed the eastern coast of the
\ac{US} north through New England, bringing large rainfall to much of the
eastern coast from North Carolina north and causing extensive inland flooding
in North Carolina (Floyd) and New England (Irene)~\citep{avila2013atlantic,
lawrence2000atlantic}. For another set of tropical cyclones (e.g., Lee in~2011
[Figure~S3], Ernesto in~2006, and Bertha in~1996), there was moderate to good
agreement for all pairwise combinations of distance, rain, and wind, but poor
agreement for all other combinations of metrics. Most of these tropical
cyclones either followed the eastern coastline north of North Carolina (Bertha
in~1996, Charley in~2004, Gaston in~2004, Barry in~2007, and Hanna in~2008), in
a pattern similar to Hurricanes Floyd in~1999 and Irene in~2011, or made
landfall in or near the Florida panhandle and cut across to exit into the
Atlantic around North Carolina (Josephine in~1996, Earl in~1998, Alberto
in~2006). However, these two sets of tropical cyclones represented the minority
of all tropical cyclones considered; for most tropical cyclones, there was low
overlap in the counties determined to be exposed to the tropical cyclones
between most pairings of metrics.

@@

When the average number of exposures per decade in each county were calculated,
strong geographical patterns were clear across tropical cyclone exposure
assessments based on these different metrics
(Figure~\ref{fig:averageexposure}). Wind-based exposure had a strong coastal
pattern, with almost all exposures in counties within about~200
\si{\kilo\metre} (124~mi) of the coastline. While exposures by rain- and
distance-based metrics were also more common in coastal areas compared to
inland areas, inland exposures were much more common based on these two metrics
compared to the wind-based metric. Rain and, to some extent, flood exposures
were characterized by a pattern defined by the Appalachian Mountains, with much
lower exposures west of the mountain range than to the east. Almost all
tornado-based exposures were in coastal states, with most in Florida, Alabama,
South Carolina, and North Carolina and almost none north of Maryland.
Flood-based exposures were highest in North Carolina and along the South
Carolina coast, as well as further north in New Jersey, southeastern
Pennsylvania, and southeastern New York. Notably, some areas of the \ac{US}
that experienced regular exposures to tropical cyclone-related excessive
rainfall and floods were not identified as areas of regular risk based on the
more commonly used tropical cyclone exposure assessment metrics of wind or
distance (Figure~\ref{fig:averageexposure}).

We conducted a case study to investigate how these differences in exposure
assessment across the five metrics considered could strongly influence
health-related studies based on physical exposure to tropical cyclones. For
this example, we compared estimates of which \ac{US} counties have the largest
expected physical exposure to tropical cyclones among a susceptible
subpopulation, Medicare beneficiaries who are reliant on electricity for
medical equipment (Figure~S6)~\citep{empower}.  We found
that a few counties (Miami-Dade County, FL; Harris County, TX) were ranked in
the top three counties of physical exposure among electricity-dependent
Medicare beneficiaries for almost all exposure metrics
(Figure~\ref{fig:topelecdependexposure}). However, there were key differences
when using different exposure metrics. For example, Florida counties dominated
the lists of top exposed counties based on wind- and tornado-based metrics,
while over half of the counties for the flood-based metric were in Mid-Atlantic
states (PA, NY).

@@

We investigated how well these data correspond with data from other available
sources (Figures~S7\,--\,11). Within the nine sample counties, storm-related
rainfall from the reanalysis precipitation dataset was well-correlated with
station-based rainfall measurements, although the reanalysis data did tend to
somewhat underestimate rainfall totals in storms with extremely high rainfall
(Figure~\ref{fig:raincomparison}). A distance constraint (county population
mean center within~500 \si{\kilo\metre} of the storm's closest approach) was
necessary for assessing storm exposure based on this metric to ensure that
rains unrelated and far from the storm track were not misattributed to storm
exposures. We found that this distance constraint was typically adequately
large to capture storm-related rain, although in rare examples of exceptionally
large storms (e.g., Hurricane Ike in 2008) or storms for which tracking was
stopped at extratropical transition (e.g., Tropical Storm Lee in 2011), some
storm-related rains were missed because of this distance constraint
(Figure~\ref{fig:rainexamples}).  The modeled wind speeds generally agreed well
with the \ac{HURDAT2} wind radii (Figure~\ref{fig:windcomparison}), with
disagreement mostly limited to large storms that were characterized by high
winds well inland (e.g., Hurricanes Ike in 2008 and Sandy in 2012;
Figure~\ref{fig:windexamples}).  Stream gage data generally indicated higher
discharge in these counties on dates identified as flood events, particularly
in counties with more reporting stream gages (e.g., Harris County, TX, with~10
reporting stream gages; Figure~\ref{fig:floodcomparison}), although these two
flooding data sources were not completely consistent.

While generally in agreement with data from other
sources, there are a few limitations. The wind data are based on modeled,
rather than observed, values, and while the modeled wind data generally agree
well with post-analysis maximum wind radii from the revised Atlantic hurricane
database \ac{HURDAT2}~\citep{landsea2013} (Figure~S7), for a few tropical cyclones
it did not (e.g., Figure~S10). The rainfall data are from re-analysis data,
which are generally well-correlated with observed ground-based station data but
sometimes oversmooth extreme measurements (Figure~S8).

### 2019-01-27 writing/ehp_tables.tex
Rearrange content

\begin{table}%[tbhp] 
\centering 
\caption{Exposure metrics considered to assess county-level exposure to 
tropical cyclones}
\begin{tabular}{p{1.5cm}p{2.5cm}p{10cm}} 
\toprule
Metric & Available & Criteria for exposure \\ \midrule 
Distance & 1988--2015 & County population mean center within 100 kilometers of storm track \\ 
Rain & 1988--2011 & County cumulative rainfall of 75 millimeters or more over the period from two days before to one day after the storm's closest approach and county population mean center within 500 kilometers of the storm track \\ 
Wind & 1988--2015 & Modeled maximum sustained wind speed at the county's population mean center 15 meters per second or higher during the storm\\ 
Flood & 1996--2015 & Flood event listed in the National Oceanic and Atmospheric Administration (NOAA) Storm Events database for the county with a start date within two days of the storm's closest approach and county population mean center within 500 kilometers of the storm track \\
Tornado & 1996--2015 & Tornado event listed NOAA Storm Events database for the county with a start date within two days of the storm's closest approach and county population mean center within 500 kilometers of the storm track\\
\bottomrule 
\end{tabular} 
\label{tab:exposuremetrics} 
\end{table}

>>>>

\begin{table}%[tbhp] 
\centering 
\caption{Exposure metrics considered to assess county-level exposure to 
tropical cyclones}
\begin{tabular}{p{1.5cm}p{2.5cm}p{10cm}} 
\toprule
Metric & Available & Criteria for exposure \\ \midrule 
Distance & 1988--2015 & County population mean center within 100 \si{\kilo\metre} of storm track \\ 
Rain & 1988--2011 & County cumulative rainfall of 75 millimeters or more over the period from two days before to one day after the storm's closest approach and county population mean center within 500 \si{\kilo\metre} of the storm track \\ 
Wind & 1988--2015 & Modeled maximum sustained wind speed at the county's population mean center 15 meters per second or higher during the storm\\ 
Flood & 1996--2015 & Flood event listed in the \ac{NOAA} Storm Events database for the county with a start date within two days of the storm's closest approach and county population mean center within 500 \si{\kilo\metre} of the storm track \\
Tornado & 1996--2015 & Tornado event listed in the \ac{NOAA} Storm Events database for the county with a start date within two days of the storm's closest approach and county population mean center within 500 \si{\kilo\metre} of the storm track\\
\bottomrule 
\end{tabular} 
\label{tab:exposuremetrics} 
\end{table}
